---
title: A glimpse into the world of Voice Leading
author: 'Lucio Cornejo'
date: '2021-06-18'
slug: a-glimpse-into-the-world-of-voice-leading
lastmod: '2021-06-18T03:54:26-05:00'
featured: no
image:
  caption: ''
  focal_point: ''
  preview_only: no
---

```{r setup, include = FALSE}
library(reticulate)
knitr::knit_engines$set(python.reticulate = reticulate::eng_python)
```

I am going to describe the surprisingly interesting musical phenomenon which 
first piqued my interest into how, from certain simple musical questions,
unexpected patterns and relationships can emerge and even be partially explained 
through mathematics. However, to make sure that non-musicians can get something 
out of this post, I'll begin describing such musical phenomenon from a mathematical perspective.

(PC use is recommended for reading this post; there are some issues on mobile)

Consider the following experiment:

1. Pick any circle.
2. Choose a ~~chord~~ finite amount of points contained in that circle. 
3. ~~Transpose~~ Rotate each of those ~~notes~~ points, along the chosen circle, 
by the same amount and in equal direction (clockwise or anticlockwise).
4. **We'll discuss this step later in the post**, but the main idea of the following graphs is that every point which has been colored corresponds to a ~~chord~~ finite set of points in the circle, which satisfies the main property that I'll explain at length later in this post.

**Now comes the good stuff ...**

```{python python-code-for-musical-graphs, echo = FALSE}
from __future__ import division
import numpy as np
import math

###### OBTENER LOS ACORDES NECESARIOS PARA EL CASO ANY FINITE CHORD SIZE ######

def ordenamiento(lista):
  # Let's assume that lista is already sorted
  # and that it has at least two elements.
  temporal = []
  for i in range(len(lista)-1):
      temporal.append([lista[0],lista[i+1]])
    
  return temporal
    
def acordes(lista,longitud_acorde):
  # Let's assume that lista is already sorted
  # and that it has at least two elements.
    
  # It holds acordes(lista,2) = ordenamiento(lista) .
    
  if longitud_acorde == 2:
      return ordenamiento(lista)    
  else:
      temporal = acordes(lista[1:],-1+longitud_acorde)
      for i in temporal:
          i.insert(0,lista[0])  
        
      for j in range(2,len(lista)-1,1):
          temp = acordes(lista[j:],-1+longitud_acorde)
          for i in temp:
              i.insert(0,lista[0])  
          temporal = temporal + temp
      return temporal
            
    
################# GRAFICAR EN BASE AL CONTRARY MOTION DEGREE #################

metric = -1    # Common musical sense of distance.
EDO = 360

def arclenght(a,b):
    temp = b%(2*np.pi)-a
    if abs(temp)>np.pi:
        if temp>0:
            temp = temp - 2*np.pi
        else:
            temp =  temp + 2*np.pi
    return temp
            
    
def dist(a,b,metric):
    if metric == -1:
        return arclenght(a,b) 
    if metric == -2:
        # Voice leading trigonometric formula
        return math.acos(np.cos(b-a))
    if metric == -3:
        # Directional arc lenght trigonometric formula
        return 2*math.atan(np.tan((b-a)/2.0))
    if metric == -4:
        # Euclidean type metrics
        #blank = (np.cos(a)-np.cos(b))**power + (np.sin(a)-np.sin(b))**power
        #return pow(blank,1.0/power)
        blank = [abs(np.cos(a)-np.cos(b)),abs(np.sin(a)-np.sin(b))]
        return max(blank)


def CPM(theta,Y,metric):  # Cyclic permutations matrix 
    m = len(Y)
    T = [(Y[i]+theta) for i in range(m)]
    GT = []
    GT.append(T)
    for i in range(1,m,1):
        temp = [T[(k+i)%m] for k in range(m)]
        GT.append(temp)
    
    mat = [[] for k in range(m)]
    for i in range(m):
        for j in range(m):
            mat[i].append(dist(Y[j],GT[i][j],metric))
            
    if abs(theta-np.pi)<0.001:
        for j in range(m):
            mat[0][j] = abs(mat[0][j])
        # This is kinda unnecesary seems the minimum
        # voice leading will probably not happen
        # with the voice motion of the first row,
        # because that's the most spread out voice leading
        # out of all the cases of the GT.
                 
    VL = [0 for i in range(m)]   
    RS = [0 for i in range(m)]
    minimal = 0
    
    for i in range(m):
        for j in range(m):
            VL[i] = VL[i] + abs(mat[i][j])
            RS[i] = RS[i] + mat[i][j]
           
        if VL[i] < VL[minimal]:
            minimal = i
            
    # Convert radians into the appropiate EDO tuning system
    rescale_1 = [i*(EDO/(2*np.pi)) for i in VL]
    rescale_2 = [i*(EDO/(2*np.pi)) for i in RS]
     
    # Contrary Motion Degree
    CMD = [rescale_1[i]-abs(rescale_2[i]) for i in range(m)]    
        
    return [rescale_1,CMD,minimal]


def graphic_data(chord_size,rotation,even_uno,even_dos,partition_precision):
  chord_size = int(chord_size) 
  partition_precision = int(partition_precision)
  even_uno = int(even_uno)
  even_dos = int(even_dos)
  temp = chord_size*partition_precision
  axis_X = [i*((2*np.pi)/temp) for i in range(0,temp,1)]
  
  rotation = rotation*(np.pi/180)
  chords = acordes(axis_X,chord_size)
  
  # List positions of most even chord of chord_size notes:
  # n*partition_precision, for n in {0,1,2, ... ,chord_size}
  edo_angle_1 = axis_X[even_uno*partition_precision]
  edo_angle_2 = axis_X[even_dos*partition_precision]
  
  if chord_size == 5:
      chords = [chords[i] for i in range(len(chords)) if edo_angle_1 in chords[i] ]
  elif chord_size == 6:
      chords = [chords[i] for i in range(len(chords)) if (edo_angle_1 in chords[i]) and (edo_angle_2 in chords[i]) ]
  
  areas = []
  area_size = 1
  results = []
  minimum_voice_leadings = []
  
  for i in range(len(chords)):
      temp = CPM(rotation,chords[i],metric)
      minimum_voice_leadings.append(temp[0][temp[2]])
      results.append(temp[1][temp[2]]) #########################################
      if abs(results[i])<0.001:
          areas.append(area_size)
      else:
          areas.append(0)
          
  # Positions related to the chords with contrary motion during such voice leading
  positions = []
  for i in range(len(areas)):
      if areas[i] == 1: 
          positions.append(i)
      
  # Relevant voice leadings
  minimum_voice_leadings = [minimum_voice_leadings[i] for i in positions]
        
  rotation = round(rotation*(180/np.pi),3) # From radians to degress
          
  if chord_size == 2:
      # Convert radians to EDO semitones
      axis_X = [ chords[i][1]*(EDO/(2*np.pi))  for i in positions ]
      axis_Y = [ 0 for i in positions ]
      axis_Z = [ 0 for i in positions ]
      
  elif chord_size == 3:
      # Convert radians to EDO semitones
      axis_X = [ chords[i][1]*(EDO/(2*np.pi))  for i in positions ]
      axis_Y = [ chords[i][2]*(EDO/(2*np.pi))  for i in positions ]
      axis_Z = [ 0 for i in positions ] # minimum_voice_leadings

  elif chord_size == 4:
      # Convert radians to EDO semitones
      axis_X = [ chords[i][1]*(EDO/(2*np.pi))  for i in positions ]
      axis_Y = [ chords[i][2]*(EDO/(2*np.pi))  for i in positions ] 
      axis_Z = [ chords[i][3]*(EDO/(2*np.pi))  for i in positions ]
  
  elif chord_size == 5:
      for i in chords:
          i.remove(edo_angle_1)
          
      axis_X = [ chords[i][1]*(EDO/(2*np.pi))  for i in positions ]
      axis_Y = [ chords[i][2]*(EDO/(2*np.pi))  for i in positions ]
      axis_Z = [ chords[i][3]*(EDO/(2*np.pi))  for i in positions ]
      
  elif chord_size == 6:
      for i in chords:
          i.remove(edo_angle_1)
          i.remove(edo_angle_2)
      
      axis_X = [ chords[i][1]*(EDO/(2*np.pi))  for i in positions ]
      axis_Y = [ chords[i][2]*(EDO/(2*np.pi))  for i in positions ]
      axis_Z = [ chords[i][3]*(EDO/(2*np.pi))  for i in positions ]
      
  return [axis_X,axis_Y,axis_Z,minimum_voice_leadings,EDO,chord_size,rotation]
        
```
```{r music-graph, echo = FALSE}
music_graph <- function(chord_size,rotation,even_uno,even_dos){
  
  partition_precision <- 0
  if(chord_size < 4){
    partition_precision <- 20
  }
  if(chord_size == 4){
    partition_precision <- 15
  }
  if(chord_size == 5){
    partition_precision <- 8
  }
  if(chord_size == 6){
    partition_precision <- 7
  }
  
  datos <- py$graphic_data(chord_size,rotation,even_uno,even_dos,partition_precision)
  x <- unlist(datos[1])  
  y <- unlist(datos[2])  
  z <- unlist(datos[3])
  edo <- unlist(datos[5])
  chord_size <- unlist(datos[6])
  rotation <- unlist(datos[7])
  
  voice_leadings <- as.numeric(unlist(datos[4]))

  library(plotly)
  datos <- data.frame(X = x, Y = y, Z = z, voice_leading_size = voice_leadings) 
  
  fig <- plot_ly(datos, x = ~X, y = ~Y, z = ~Z, opacity = 0.7, mode = "markers",
                 text = ~paste0("Minimal voice leading size: ", round(voice_leading_size,3),
                                " degrees"),
                        hoverinfo = text,
               marker = list(
                        size = 2.5, color = ~voice_leading_size,
                        colorscale = list(c(0,1),c("rgb(105, 149, 237)",
                                                   "rgb(255, 34, 34)")),
                        showcase = TRUE
                        )
                )
  fig <- fig %>% add_markers()
  fig <- fig %>% layout(title = paste("\n Chord size: ",chord_size,
                                "      Rotation: ", round(rotation,3), " degrees"),
            scene = list(xaxis = list(title = 'X',range = c(0,edo) ),
           yaxis = list(title = 'Y',range = c(0,edo) ), zaxis = list(title = 'Z',
           range = c(0,edo)), camera = list(center = list(x = 0,y = 0,z = 0),
           eye = list(x = 1.25,y = 1.25,z = 1.25),
           up = list(x = 2.5,y = 0,z = 1))),
           annotations = list(x = 1.0, y = 1.0, text = '', showarrow = FALSE )
           )
  fig
}
```
```{r graphics, message = FALSE, echo = TRUE}
# Display interactive graph 1
music_graph(2,150,0,0)

# Display interactive graph 2
music_graph(3,90,0,0)

# Display interactive graph 3
music_graph(4,120,0,0)
``` 
\

Reader: *Ok, so what the hell was that?*

Author: *Yeah ... sorry for the spoilers. Now I'll provide the necessary background to understand step 4 and the graphs shown.*

\

## Visual examples

From a musical perspective, we'll be dealing with **chords** (finite sets
of at least two distinct notes) and their **transpositions** (translating them
by some distance), but, from a mathematical perspective, we'll be working with
points in a circle (the border, not the inside) and rotating them by some angle.

For simplicity, we'll work with the unitary circle, the circle with radius 1 
and whose center is the origin of the Cartesian Plane.
Each point in the unitary circle can be thought of as a possible musical note,
if we conveniently match them.  

In that sense, let's work with the following pairing:

<!--
Creating a table using HTML 
-->

<table>
  <tr>
    <th> Note </th>
    <td> C  </td>
    <td> Db </td>
    <td> D  </td>
    <td> Eb </td>
    <td> E  </td>
    <td> F  </td>
    <td> Gb </td>
    <td> G  </td>
    <td> Ab </td>
    <td> A  </td>
    <td> Bb </td>
    <td> B  </td>
  </tr>
  <tr>
    <th> Angle </th>
    <td> $ (0/12)*360^{°} $ </td>
    <td> $ (1/12)*360^{°} $ </td>
    <td> $ (2/12)*360^{°} $ </td>
    <td> $ (3/12)*360^{°} $ </td>
    <td> $ (4/12)*360^{°} $ </td>
    <td> $ (5/12)*360^{°} $ </td>
    <td> $ (6/12)*360^{°} $ </td>
    <td> $ (7/12)*360^{°} $ </td>
    <td> $ (8/12)*360^{°} $ </td>
    <td> $ (9/12)*360^{°} $ </td>
    <td> $ (10/12)*360^{°}$ </td>
    <td> $ (11/12)*360^{°}$ </td>
  </tr>
</table>

\ 

Let's start with a simple case: Rotating the chord "C major" by \(90^{°}\; .\)

Such chord is comprised of the notes C, E and G; notes which we can represent 
in the Cartesian Plane (given the note \( \equiv \) angle pairing
we have already established) as the following points:

C : \( \left(\cos(0^{°}),\sin(0^{°})\right) \)

E : \( \left(\cos(120^{°}),\sin(120^{°})\right) \)

G : \( \left(\cos(210^{°}),\sin(210^{°})\right) \)

However, it's much simpler to represent chords not by the coordinates of the 
points which comprise such chord, but by the unique angles,
inside \([\,0^{°};360^{°}\,[\;\), associated to those points.\
In that sense, let's work with the following equivalence: \
C major \(\equiv\) {C; E; G} \(\equiv (0,120,210) \; ,\) \
where the angles in the **vector representation** of the chord will be written
in degrees and from lowest to highest angle.

Now, let's rotate each of the points which comprise C major by \( 90^{°} \) 
with respect to the origin, as we can see in the following video, 
where the initial points shown (in green) correspond to the notes 
C, E and G, in that order.

<video width = 100% controls>
  <source src = "acorde_0.mp4" type = "video/mp4">
</video>

From a musical perspective, we've **transposed** the C major chord by \
\( 90^{°} = \frac{360^°}{4} \equiv \frac{12 \text{ semitones}}{4} = 
3 \text{ semitones }\)
and landed on Eb major, but, mathematically, we can think of having 
**rotated** the initial chord.

The **set of trajectories** that the notes/points from C major take 
*along the unitary circle* to get to the notes in Eb major is called the 
**voice leading** between C major and Eb major. 

Let me clarify that the voice leadings which we'll be dealing with in this post are precisely those which satisfy that 
**no pair of distinct notes from the initial chord end up at the same note in the transposed chord.** 
In that sense, when we tranpose/rotate any chord made up of \( n \) notes, 
the number of possible voice leadings between such chord and its transposition
is precisely 
\( n! = n\cdot(n-1)\cdot(\dots)\cdot 2 \cdot 1 \enspace\).
<br> That's because every note \(x\) in the initial chord is moving (in a voice
leading) to a note \( y \) in the transposed chord, unique to \(x\).
That means that no other note from the initial chord is also moving to \( y\,\); 
therefore, combinatorics is all we need. \
**Curious fact:** We actually only need to check \(n\) special kinds of voice
leadings, not \( n! \), in order to find one which is, at least, a minimal voice
leading.
The justification comes from a mathematical theorem which we'll explore further 
in the next post, where it will play a **key role**. If you are interested,
you can read about it in this [article](https://www.jstor.org/stable/25171405) .

\

Going back to our concrete case, let's see some voice leading examples for the 
transposition from C major to Eb major in the following video:

<video width = 100% controls>
  <source src = "acorde_1.mp4" type = "video/mp4">
</video>

Notice from the video that the trajectories which define the voice leading correspond 
to the shortest of the two paths that a point could have taken to get to the other.
In the case where both paths have the same length, that is, for a \( 180^{°} \)
rotation, we'll choose, as a convention, the anticlockwise trajectory
to be a part of the voice leading.

In this post, we'll focus on two properties that we'd like a voice leading (*VL*)
between a chord and some of its transpositions to have:

1. The sum of the **length of the trajectories** which define *VL*, also known as the **voice leading size**, is the **smallest one** out of all the possible voice leadings between the initial chord and its transposition. <br> Any voice leading which satisfies this property is said to be a **minimal voice leading**.
2. The trajectories which define *VL* share the same direction, that is, they are 
**all clockwise** or **all anticlockwise**. <br> Any voice leading which satisfies this 
property is said to have **no contrary motion**.

Partly, what I've shown you in the initial graphs is which chords, of some fixed 
chord size, satisfy that after transposing/rotating them by a certain amount,
there exists a voice leading, between the initial and transposed chord, that is
a *minimal voice leading* and *has no contrary motion*. If some chord satisfies 
such property, we'll say that it has an **unidirectional minimal voice leading**.

I wouldn't blame you if you were just thinking the following:
<br> "Why would a minimal voice leading be unidirectional?.
<br> The properties 1. and 2. do **NOT** seem related at all".
<br> You'd be partly right, in fact, because the property 1. deals with 
minimizing trajectories' lenghts, while the property 2. refers only to such 
trajectories' directions.

However, both properties happen to be fairly related, and even extremely related
when we focus on \( 180^{°} \) rotations of chords, a situation where symmetry 
arises with respect to the possible voice leadings. But, such connection will
be explored in the next post because that's a deep subject as well.

\

Let's get back to our example, rotating the C major by \( 90^{°} \).
<br> Does it exist an unidirectional minimal voice leading in that case?

After checking by hand the \( 3! = 6 \,\) possible voice leadings, 
we actually find one unidirectional minimal voice leading, the one that I
labeled as *voice leading 3*. 
Such voice leading is comprised of *clockwise trajectories* and it has a voice 
leading size of \( 90^{°}\; . \) \
Of course, there is some ambiguity with respect to the direction of the 
trajectory in *voice leading 3* which has a size of \(0^{°}\); but, we can 
ignore its direction (in the same way that the number zero has no sign)
when we are checking if some voice leading is a unidirectional minimal 
voice leading. The other trajectories in *voice leading 3* have the same
direction, so that's all that matters to us. \
Notice that *voice leading 1*  and *voice leading 2* fail to be minimal
voice leadings because their respective voice leading sizes,
\( 90^{°} \cdot 3 = 180^{°}\;
\text{ and }
\: 150^{°} + 180^{°} + 120^{°} = 450^{°}  \), are greater than the voice leading
size of the minimal voice leading that I already presented (*voice leading 3*).\
Also note that *voice leading 1* has no contrary motion but *voice leading 2*
does have contrary motion.

\

## Putting it all together

We are 80% ready to understand the graphs shown previously.

The **first step** is to **choose a chord size** (\(n\)), that is,
the amount of distinct points in a circle that we will be rotating afterwards.

The **second step** is to **chose some rotation** ('rotation'), a value between
\( 0^{°} \) and \(360^{°} \), that we will apply to *'all'* chords with 
the size already chosen.

In that sense, if we use the vector representation of chords via their angles,
as we did with C major '=' \( (0^{°},120^{°},210^{°}) \), the domain of the 
function, *music_graph(...)*, function from which these graphs shown come from,
would be 
\[
\left\{\,
(x_0,x_1,x_2, \dots ,x_{n-1}) \in \mathbb{R}^{n} \, : \,\
0 \leq x_0 < x_1 < \dots < x_{n-1} < 360
\,\right\}
\]

However, we can be more efficient and work only with a special subset of such 
domain, which would be the new domain of *music_graph(..)*, yet still achieve 
our goal of finding for which chords of size \(n\) there exists an unidirectional
minimal voice leading when they are rotated by 'rotation'.
Such efficiency is a direct result of the following:


**Important property: ** If there exists a unidirectional minimal voice leading
with respect to some rotation \(\theta\) of some chord \(X\), then there also
exists a unidirectional minimal voice leading with respect to the rotation
\(\theta\) of any chord which is itself some rotation of \(X\).

As a consequence, based on the fact that for the rotation \( 60^{°} \) of the
C major chord, which we can label via its angles as \( (0^{°},120^{°},210^{°})\),
there exists a unidirectional minimal voice leading, if we were to rotate also by
\(60^{°}\) chords whose angles are of the form
\( ( (0+\alpha)^{°},(120+\alpha)^{°},(210+\alpha)^{°})\), where \(\alpha\) can
be any real number, then there would also exist at least one unidirectional 
minimal voice leading for those chords.

This means that the domain of *music_graph(...)* can be reduced in one dimension,
after which it would take the following form as the new domain of 
*music_graph(...)*:

\[\left\{\,
(0,x_1,x_2, \dots ,x_{n-1})\in \mathbb{R}^{n} \, : \,
0 < x_1 < \dots < x_{n-1} < 360
\,\right\}\,\,\dots\,(\triangle)\,\,\]

The **third step** is to check, for every chord 
\((0,x_1,x_2, \dots ,x_{n-1})\,\) in \((\triangle)\)
if there exists an unidirectional minimal voice leading after rotating such
chord by 'rotation'. 
<br> Practically, we can't check each of those chords, because
there's an infinite amount of them, but the computer will check a sufficient 
amount so that we can accurately approximate how the graph of *music_graph(...)*
actually looks like.
<br> Let's denote the set of chords in \((\triangle)\) for which 
there exists an unidirectional minimal voice leading after rotating those chords
by 'rotation', as  *'success'*.

The **fourth step** is displaying, **appropriately**, on the 3-dimensional space 
\(\mathbb{R}^3\,\) the elements of *'success'*. 
Allow me to explain the main 4 cases:

Case \(n = 2\): 
<br> The domain in \((\triangle)\) becomes
\(\left\{\,
(0,x)\in\mathbb{R^2}\, : \,
0 < x < 360
\,\right\}\,.\)
<br> The elements of *'success'* are special chords represented by 
\( (0,\bar{x})\), with \(0<\bar{x}<360\). 
Such chords will be displayed to be in *'success'* via coloring the points
\((\bar{x},0,0) \in \mathbb{R}^3\,\,.\)
The last two zeros in \((\bar{x},0,0) \) I added them in for simplicity only,
they aren't important.

Case \(n = 3\): 
<br> The domain in \((\triangle)\) becomes
\(\left\{\,
(0,x,y)\in\mathbb{R^3}\, : \,
0 < x  < y < 360
\,\right\}\,.\)
<br> The elements of *'success'* are special chords represented by 
\( (0,\bar{x},\bar{y})\), with \(0<\bar{x}<\bar{y}<360\). 
Such chords will be displayed to be in *'success'* via coloring the points
\((\bar{x},\bar{y},0) \in \mathbb{R}^3\,\,.\)
The last zero in \((\bar{x},\bar{y},0) \) I added in for simplicity only,
it's not important.

Case \(n = 4\): 
<br> The domain in \((\triangle)\) becomes
\(\left\{\,
(0,x,y,z)\in\mathbb{R^4}\, : \,
0 < x  < y < z < 360
\,\right\}\,.\)
<br> The elements of *'success'* are special chords represented by 
\( (0,\bar{x},\bar{y},\bar{z})\), with \(0<\bar{x}<\bar{y}<\bar{z}<360\). 
Such chords will be displayed to be in *'success'* via coloring the points
\((\bar{x},\bar{y},\bar{z}) \in \mathbb{R}^3\,\,.\)

Case \(4 < n \leq 6\):
<br> We've runned out of dimensions, because now the graphs are either
4 dimensional or 5 dimensional. 
For that reason, in order to be able to visualize the graph 
in a 3 dimensional space, \( n-4 \) of the variables in 
\( (0,x_1,x_2, \dots ,x_{n-1}) \) will need to be chosen in advance and
be kept fixed during the **third step**.
<br> Which angles will remain fixed is what the last two numbers in 
the command *music_graph(...)* refer to.

\

How to interpret the function 
music_graph(chord_size, rotation, \(m_1\), \(m_2\)) :
<br> We already know what *chord_size* and *rotation* refer to.
<br> The extra caveat to consider is that the chords represented in 
\((\triangle)\) must contain the angles 
\( \frac{m_1}{\text{chord_size}}\cdot 360^{°}\,\, \) and
\(\,\,\frac{m_2}{\text{chord_size}}\cdot 360^{°}\,\) . 

Now you can see why in the case \(m_1 = m_2 = 0\) we didn't need to fix
any angle in \((\triangle)\), because the angle \( \,0^{°}\) is already present
in those chords which we are considering.

The **fifth step**, the final one, is to color appropriately the points
displayed in the **fourth step**. 
<br> After having chosen the parameters in 
*music_graph(chord_size, rotation, \(m_1\), \(m_2\))* and doing all the
first 4 steps just mentioned, we established a finite set of points 
to represent/approximate \((\triangle)\), after which we selected a subset of it,
let's call it *'success_approximation'*, which represents/approximates *'success'*.
Each element of the finite set *'success_approximation'* has a certain minimal 
voice leading size, so, I'll define a coloring based on such 
minimal voice leading sizes. 
<br> **Blue** will be the color of the displayed point
(**fourth step**) for which its associated chord had the *smallest* minimal
voice leading size (min_VL) with respect to the minimal voice leading sizes of all 
the chords in *'success_approximation'*.
<br> **Red** will be the color of the displayed point
(**fourth step**) for which its associated chord had the *biggest* minimal
voice leading size (max_VL) with respect to the minimal voice leading sizes of all 
the chords in *'success_approximation'*. 
<br> The other points displayed in the **fourth step** will be colored with
a color between blue and red, depending on the position of their minimal voice
leading size in the interval \( [\,\text{ min_VL };\text{ max_VL }\,]\,\, . \)

Notice that there may be more than one chord in *'success_approximation'*
with the smallest (or biggest) minimal voice leading size, out of the minimal
voice leading sizes of all the chords in *'success_approximation'*.
<br> Besides this, it may also be the case that, for certain parameters in 
*music_graph(chord_size, rotation, \(m_1\), \(m_2\))*, all the chords in
*'success_approximation' * have precisely the same minimal voice leading size.
In that case, all the displayed points (**fourth step**) would be colored 
in **purple**.

\

Of course, we don't have to stop at the case of chords with size 6,
we could consider, for example, chords of size 666 and so on.
However, I only coded the function *music_graph(...)* till chord size 6,
so that's the limit for now.

I encourage you to play around with the graphics in this post. 
For example, notice the bizarre fact that each of the graphs shown 
contains regions where the minimal voice leading size remains constant, 
that is, where the coloring applied doesn't change.

\

In conclusion, I hope that after reading this post, you've gained some 
appreciation for music. We only focused on rotating points along a circle,
while checking for certain properties of the trajectories of the points,
a very simple problem indeed; yet, the graphs shown displayed no 
simplicity whatsoever. Do we even know which mathematical equation determines 
the shape of the graphs? From what I've researched online, 
it's not known publicly.
<br> I **do** know that precise mathematical equation, but I'd rather not share
it till I present it in my undergraduate thesis.

\

**Thanks for reading!**
<br> This post took a huge amount of work, so I am glad you made it till here.
<br> I'll implement an online graph calculator (probably in early August) so that 
you can play around changing the inputs of the function *music_graph(...)* 
and see what bizarre shapes come into existence.

For now, I can leave you with some more graph examples in case you are curious
about how other cases look like:

<!--
## Randomly generated graphs for the curious mind

I'll now include five graphs which change everytime you reload the page,
so that you can play around with more examples and hopefully understand 
the phenomenon described in this post much better.
-->

```{r extra-graphics, message = FALSE, echo = TRUE}
music_graph(2,90,0,0)
music_graph(2,180,0,0)
music_graph(3,65,0,0)
music_graph(3,180,0,0)
music_graph(4,155,0,0)
music_graph(4,180,0,0)
music_graph(5,55,2,0)
music_graph(6,35,2,3)
music_graph(6,180,3,4)
```